Liveness-Based Garbage Collection for Lazy Languages

### 2015-11-24 paper/paper.tex
Updated paper with latest example

\begin{figure*}[t!]
\renewcommand{\arraystretch}{1}
  \begin{tabular}{cc}
    \begin{minipage}{.40\textwidth}
        \small
        \renewcommand{\arraystretch}{1}{
          \begin{uprogram}
            \UNL{1} (\DEFINE\ (\length~\xl)
            \UNL{2}  $\pi_1\!\!:\, $(\LET\ \px\ $\leftarrow $\
(\NULLQ~\xl) \IN
            \UNL{3} \hspace*{.05cm} $\pi_2\!\!:\,$(\SIF\
$\psi_1\!\!:\,$ \px
            \UNL{4} \hspace*{.27cm} $\pi_3\!\!:\,
            $(\LET\ \pv\ $\leftarrow 0$ \IN
            \UNL{5} \hspace*{.32cm} $\pi_4\!\!:\,
(\SRETURN~\psi_2:\pv)$
            \UNL{4} \hspace*{.29cm}    $\pi_5\!\!:\, $(\LET~\pu\
$\leftarrow$  (\CDR~\xl)  \IN
            \UNL{5} \hspace*{.34cm}   $\pi_6\!\!:\, $(\LET~\py\
$\leftarrow$  (\length~\pu)  \IN
            \UNL{6} \hspace*{.34cm} $\pi_7\!\!:\,
            $(\LET~\pz\ $\leftarrow$ (1~+~\py) \IN
            \UNL{7} \hspace*{.34cm} $\pi_8\!\!:\,
(\SRETURN~\psi_3:\pz)$)))))))
        \end{uprogram}}
        \renewcommand{\arraystretch}{1}{
	  \begin{uprogram}
	  \UNL{1} $(\DEFINE\ (\main)$
           \UNL{2} \!\!$\pi_9\!\!:\, (\LET\  \pa\  \leftarrow$
           (
           \scalebox{0.8}{\psframebox[framearc=.5,
	fillcolor=lightgray,fillstyle=solid,framesep=2pt]{%
             \begin{tabular}{@{}c@{}}
               {a BIG closure}
             \end{tabular}}}
           ) \IN  
          \UNL{3} \!\!$\pi_{10}\!\!:\, (\LET\  \pb\  \leftarrow ($+$\ \pa\ \acdr)$
\IN
	  \UNL{4}   \hspace*{.05cm}$\pi_{11}\!\!:\,      $ (\LET\ \pc\
$\leftarrow  (\CONS\ \pb\ \NIL)$ \IN
          \UNL{5}   \hspace*{.15cm}    $\pi_{12}\!\!:\,
          $(\LET\ \pw\  $\leftarrow  (\length\ \pc)$ \IN
          \UNL{6}  \hspace*{.25cm}  $\pi_{13}\!\!:\,
(\SRETURN~\psi_4:\pw)))))$
\end{uprogram}}
        %%}
    \end{minipage}

@@

    \begin{minipage}{.51\textwidth}

@@

Given  a  function  $\mathit{f}$,  we now  describe  how  to  generate
equations  for the  demand  transformation \Lfonly$_\mathit{f}$.   The
program in Figure~\ref{fig:mot-example2} serves  as a running example.
Starting   with    a   symbolic   demand   $\sigma$,    we   determine
\Lfun{e_{\mathit{f}}}{\sigma}{\Lfonly}.   In  particular, we  consider
$\Lanv{1}{x_i}$,  the liveness  of  the $i^{\mbox{\footnotesize  th}}$
parameter   $x_i$  at   the  program   point  at   the  beginning   of
$e_{\mathit{f}}$   (assumed  to   be  $\pi_1$).   By  the   rule  {\sc
  live-define},      this     should      be      the     same      as
$\Lf{f}{i}{\sigma}$. Applying this to \length, we have:
\begin{eqnarray*}
&& \Lanv{1}{\xl} = \clazy\sigma \cup \acdr\Lf{\length}{1}{\clazy\sigma}
  \cup \clazy\Lf{\length}{1}{\clazy\sigma}
\end{eqnarray*}
and  the only equation defining \Lfone{\length} is:
  \begin{eqnarray*}
   && \Lf{\length}{1}{\sigma}
    =  \clazy\sigma \cup \acdr\Lf{\length}{1}{\clazy\sigma} \cup
\clazy\Lf{\length}{1}{\clazy\sigma}
\end{eqnarray*}

In general, the equations for  \Lfonly\ are recursive since, as in the
case  of  \length,  \Lan{1}\  may   have  been  defined  in  terms  of
\Lfone{\mathit{f}}.  However,  it is desirable  to have a  closed form
solution for $\Lfone{\mathit{f}}$.  As mentioned in \cite{asati14lgc},
each  of the liveness  rules modifies  a demand  by prefixing  it with
symbols  in the  alphabet  $\lbrace \acar,  \acdr,\bcar, \bcdr,  \clazy
\rbrace$, and therefore we can assume that $\Lf{f}{i}{\sigma}$ has the
closed form:
\begin{eqnarray}
\label{eq:LF:DI}
  \Lf{f}{i}{\sigma} = \Df{f}{i}\sigma
\end{eqnarray}
where \Df{f}{i} are sets of strings over the alphabet mentioned above.
Substituting   the   guessed   form   in   the   equation   describing
\Lfonly$_{\mathit f}$, and factoring  out $\sigma$, we get an equation
for  \Df{f}{i} that  is  independent of  $\sigma$.   Any solution  for
\Df{f}{i}  yields a  solution for  \Lfonly$_{\mathit f}$.   Applied to
\Lfonly$_{\mathit \length}$, we get:
  \begin{eqnarray*}
&&  \Lf{\length }{1}{\sigma} = \Df{\length}{1}\sigma,~\mbox{and}\\
&&   \Df{\length}{1} = \clazy \cup \acdr\Df{\length}{1}\clazy
       \cup \clazy\Df{\length}{1}{\clazy}
  \end{eqnarray*}

Note that this equation can also be viewed as a CFG with \{\acdr,
\clazy\} as terminal symbols and \Df{\length}{1} as the sole
non-terminal.

@@

For  the  running  example,   $\length$  has  calls  from  $\main$  at
$\pi_{12}$ and a recursive call at $\pi_6$.
So $\sigma_{\length} =
     \deltacall{\length}{\pi_{12}}{\main}  \cup
\deltacall{\length}{\pi_6}{\length}$.
Filling in  the values gives:
\begin{eqnarray*}
\sigma_{\length}    &=&
 \{\epsilon\}  ~\cup~{\clazy\sigma_{\mathit{\length}}}
\end{eqnarray*}
As examples, the liveness of \Lanv{1}{\pl} and \Lanv{9}{\pa} in terms
of  $\sigma_{\length}$ are:
\begin{align*}
\Lanv{1}{\pl} &= (\clazy \cup \acdr\Df{\length}{1}\clazy
  \cup \clazy\Df{\length}{1}{\clazy})\sigma_{\mathit{\length}}
 \\
\Lanv{9}{\pa} &= \clazy \bcar \Df{\length}{1}\{\epsilon\}
\end{align*}

>>>>

\begin{figure*}[t!]
\renewcommand{\arraystretch}{1}
  \begin{tabular}{cc}
    \begin{minipage}{.50\textwidth}
        \small
        \renewcommand{\arraystretch}{1}{
          \begin{uprogram}
            \UNL{1} (\DEFINE\ (\fun~\px\ \pn)
            \UNL{2}  $\pi_1\!\!:\, $(\LET\ \py\ $\leftarrow $\
(\length~\px) \IN
            \UNL{3} $\pi_2\!\!:\, $(\LET\ \pl\ $\leftarrow $\
(\makelist~\pn) \IN
            \UNL{4} \hspace*{.05cm} $\pi_3\!\!:\,$(\SIF\
$\psi_1\!\!:\,$ \py
            \UNL{5} \hspace*{.27cm} $\pi_4\!\!:\, $(\LET\ \xhd\ $\leftarrow$\ (\CAR\ \pl) \IN
            \UNL{6} \hspace*{.27cm} $\pi_5\!\!:\, $(\LET\ \xtl\ $\leftarrow$\ (\CDR\ \px) \IN
            \UNL{7} \hspace*{.27cm} $\pi_6\!\!:\, $(\LET\ \rest\ $\leftarrow$\ (\fun~ \xtl\ \xhd) \IN
            \UNL{8} \hspace*{.27cm} $\pi_7\!\!:\, $(\LET\ \ret\ $\leftarrow$\ (\CONS\ \py\ \rest) \IN
            \UNL{9} \hspace*{.32cm} $\pi_8\!\!:\,
(\SRETURN~\psi_2:\ret)))))$
            \UNL{5} \hspace*{.29cm}    $\pi_9\!\!:\, $(\LET~\ps\
$\leftarrow$  (\Sum~\pl)  \IN
            \UNL{6} \hspace*{.34cm}   $\pi_{10}\!\!:\, $(\LET~\pu\
$\leftarrow$  (+~1~\py)  \IN
            \UNL{7} \hspace*{.34cm} $\pi_{11}\!\!:\,
            $(\LET~\pr\ $\leftarrow$ ($/$~\ps~\pu) \IN
            \UNL{8} \hspace*{.27cm} $\pi_{12}\!\!:\, $(\LET\ \res\ $\leftarrow$\ (\CONS\ \pr\ \NIL) \IN
            \UNL{9} \hspace*{.34cm} $\pi_{13}\!\!:\,
(\SRETURN~\psi_3:\res)$)))))))))
        \end{uprogram}}
        \renewcommand{\arraystretch}{1}{
	  \begin{uprogram}
	  \UNL{1} $(\DEFINE\ (\main)$
           \UNL{2} \!\!$\pi_{14}\!\!:\, (\LET\  \pa\  \leftarrow$
           (\CONS\ $1$\ (\CONS\ $2$\ \NIL) \IN  
          \UNL{3} \!\!$\pi_{15}\!\!:\, (\LET\  \pb\  \leftarrow $5$)$
\IN
	  \UNL{4}   \hspace*{.05cm}$\pi_{16}\!\!:\,      $ (\LET\ \pc\
$\leftarrow  (\fun\ \pa\ \pb)$ \IN
          \UNL{5}   \hspace*{.15cm}    $\pi_{17}\!\!:\,
          (\SRETURN~\psi_4:\pc)))))$
\end{uprogram}}
        %%}
    \end{minipage}

@@

    \begin{minipage}{.48\textwidth}

@@

Given  a  function  $\mathit{f}$,  we now  describe  how  to  generate
equations  for the  demand  transformation \Lfonly$_\mathit{f}$.   The
program in Figure~\ref{fig:mot-example2} serves  as a running example.
Starting   with    a   symbolic   demand   $\sigma$,    we   determine
\Lfun{e_{\mathit{f}}}{\sigma}{\Lfonly}.   In  particular, we  consider
$\Lanv{1}{x_i}$,  the liveness  of  the $i^{\mbox{\footnotesize  th}}$
parameter   $x_i$  at   the  program   point  at   the  beginning   of
$e_{\mathit{f}}$   (assumed  to   be  $\pi_1$).   By  the   rule  {\sc
  live-define},      this     should      be      the     same      as
$\Lf{f}{i}{\sigma}$. Applying this to \length, we have:

\begin{eqnarray*}
&& \Lanv{1}{\xl} = \clazy\sigma \cup \acdr\Lf{\length}{1}{\clazy\sigma}
  \cup \clazy\Lf{\length}{1}{\clazy\sigma}
\end{eqnarray*}
and  the equation defining \Lfone{\fun} is:
  \begin{eqnarray*}
   && \Lf{\fun}{1}{\sigma}
    =  \clazy\sigma \cup \Lf{\length}{1}{\bcar\sigma_{\mathit{\length}}} \cup  \Lf{\length}{1}{\clazy\sigma} 
       \cup  \acdr\Lf{\fun}{1}{\bcdr\sigma_{\mathit{\fun}}}
\end{eqnarray*}

In general, the equations for  \Lfonly\ are recursive since, as in the
case  of  \fun,  \Lan{1}\  may   have  been  defined  in  terms  of
\Lfone{\mathit{f}}.  However,  it is desirable  to have a  closed form
solution for $\Lfone{\mathit{f}}$.  As mentioned in \cite{asati14lgc},
each  of the liveness  rules modifies  a demand  by prefixing  it with
symbols  in the  alphabet  $\lbrace \acar,  \acdr,\bcar, \bcdr,  \clazy
\rbrace$, and therefore we can assume that $\Lf{f}{i}{\sigma}$ has the
closed form:
\begin{eqnarray}
\label{eq:LF:DI}
  \Lf{f}{i}{\sigma} = \Df{f}{i}\sigma
\end{eqnarray}
where \Df{f}{i} are sets of strings over the alphabet mentioned above.
Substituting   the   guessed   form   in   the   equation   describing
\Lfonly$_{\mathit f}$, and factoring  out $\sigma$, we get an equation
for  \Df{f}{i} that  is  independent of  $\sigma$.   Any solution  for
\Df{f}{i}  yields a  solution for  \Lfonly$_{\mathit f}$.   Applied to
\Lfonly$_{\mathit \fun}$, we get:
  \begin{eqnarray*}
&&  \Lf{\fun }{1}{\sigma} = \Df{\fun}{1}\sigma,~\mbox{and}\\
&&   \Df{\fun}{1} =  \clazy \cup \Df{\length}{1}\bcar \cup  \Df{\length}{1}\clazy 
       \cup  \acdr\Df{\fun}{1}\bcdr
  \end{eqnarray*}

Note that this equation can also be viewed as a CFG with \{\acdr, \bcar, \bcdr,
\clazy\} as terminal symbols and \Df{\length}{1} and \Df{\fun}{1} as the 
non-terminals.

@@

For  the  running  example,   $\fun$  has  calls  from  $\main$  at
$\pi_{16}$ and a recursive call at $\pi_6$.
So $\sigma_{\length} =
     \deltacall{\fun}{\pi_{16}}{\main}  \cup
\deltacall{\fun}{\pi_6}{\length}$.
Filling in  the values gives:
\begin{eqnarray*}
\sigma_{\fun}    &=&
 \{\sigma_{\mathit{all}}\}  ~\cup~{\bcdr\sigma_{\mathit{\fun}}}
\end{eqnarray*}
As examples, the liveness of \Lanv{2}{\pl} and \Lanv{1}{\py} in terms
of  $\sigma_{\fun}$ are:
\begin{align*}
\Lanv{2}{\pl} &= \clazy \cup \acar\Lf{\fun}{2}{\bcdr\sigma_{\mathit{\fun}}}
 \\
\Lanv{1}{\py} &= \clazy \cup \bcar\sigma_{\mathit{\fun}}
\end{align*}
